\textit{Supplemental materials for:} High-order cognition is supported by information-rich but compressible brain activity patterns

### 2023-03-16 paper/main.tex
checking in edits

We examined fMRI data collected as participants listened to an auditory
recording of a story, scrambled recordings of the story, or underwent a resting
state scan. We found that cognitively richer stimuli evoked more reliable
(i.e., consistent across people) and information rich brain activity patterns.
The brain patterns evoked by cognitively richer stimuli were also more
compressible, in that each individual component provided more ``signal'' to
temporal decoders relative to components of data from less cognitively rich
conditions (Fig.~\ref{fig:inflection}). Over time (e.g., as the experiment
progressed), these phenomena were strengthened. Specifically, across story
segments, data from more cognitively rich conditions became more informative
and compressible, and data from less cognitively rich conditions became
\textit{less} informative and compressible (Fig.~\ref{fig:inflection-thirds}).
We also repeated these analyses separately for different brain networks. We
found that networks traditionally associated with higher-level cognitive
functions tended to provide more informative brain patterns than networks
traditionally associated with lower level cognitive functions
(Fig.~\ref{fig:networks}). Finally, we examined the most dominant components of
the brain activity patterns from each experimental condition. We used a reverse
inference approach~\citep{RubiEtal17} to identify the terms in the neuroimaging
literature most commonly associated with the corresponding maps. As summarized
in Figure~\ref{fig:neurosynth-summary}, we found that terms associated with memory and
sensory processing were associated with the strongest components in all three
story listening conditions. Terms associated with sensory integration were
associated with the strongest components in the intact and paragraph-scrambled
conditions. Terms associated with sentence comprehension, emotion, and valence
were associated with the strongest components in the intact condition. Finally,
terms associated with the default mode network were associated with the
strongest components in the word-scrambled and resting state conditions. Taken
together, our findings indicate that the informativeness and compressibility of
our brain activity patterns are task-dependent, and these properties change
systematically with factors like cognitive richness and depth of processing.

@@

We used an across-participant correlation-based classifier to decode which
stimulus timepoint matched each timepoint's neural pattern. For a given value
of $k$ (i.e., number of principal components), we first used group PCA to
project the data from each condition into a shared $k$-dimensional space. Next,
we divided the participants into two groups: a template group,
$\mathcal{G}_{\mathrm{template}}$ (i.e., training data), and a to-be-decoded
group, $\mathcal{G}_{\mathrm{decode}}$ (i.e., test data). We averaged the
projected data within each group to obtain a single $T$ by $k$ matrix for each
group. Next, we correlated the rows of the two averaged matrices to form a $T$
by $T$ decoding matrix, $\mathbf{\Lambda}$. In this way, the rows of
$\mathbf{\Lambda}$ reflected timepoints from the template group, while the
columns reflected timepoints from the to-be-decoded group. We used
$\mathbf{\Lambda}$ to assign temporal labels to each timepoint (row) from the
test group's matrix, using the row of the training group's matrix with which it
was most highly correlated. We repeated this decoding procedure, but using
$\mathcal{G}_{\mathrm{decode}}$ as the template group and
$\mathcal{G}_{\mathrm{template}}$ as the to-be-decoded group. Given the true
timepoint labels (for each group), we defined the decoding accuracy as the
average proportion of correctly decoded timepoints, across both groups (where
chance perfomance is $\frac{1}{T}$). In Figures~\ref{fig:inflection}
and~\ref{fig:inflection-thirds} we report the decoding accuracy for each
condition and value of $k$, averaged across $n = 100$ cross validation folds.

\textbf{JRM NOTE: UPDATE THE NEXT TWO SUB-SECTIONS}
\subsection*{Reverse inference}

To help interpret the brain activity patterns we found within the contexts of
other studies, we created summary maps of each principal component, for each
experimental condition, by summing together the 20 HTFA-derived RBF nodes (see
\textit{Hierarchical Topographic Factor Analysis}) with the highest absolute
value weights for each of the top 5 components (Figs.~\ref{fig:neurosynth-pca},
\synthThirds). We then carried out a meta analysis using
Neurosynth~\citep{RubiEtal17} to identify the 5 terms most commonly associated
with the given map.

@@

>>>>

We examined fMRI data collected as participants listened to an auditory
recording of a story, scrambled recordings of the story, or underwent a resting
state scan. We found that cognitively richer stimuli evoked more reliable
(i.e., consistent across people) and information rich brain activity patterns.
The brain patterns evoked by cognitively richer stimuli were also more
compressible, in that each individual component provided more ``signal'' to
temporal decoders relative to components of data from less cognitively rich
conditions (Fig.~\ref{fig:inflection}). Over time (e.g., as the experiment
progressed), these phenomena were strengthened. Specifically, across story
segments, data from more cognitively rich conditions became more informative
and compressible, and data from less cognitively rich conditions became
\textit{less} informative and compressible (Fig.~\ref{fig:inflection-quarters}).
We also repeated these analyses separately for different brain networks. We
found that networks traditionally associated with higher-level cognitive
functions tended to provide more informative brain patterns than networks
traditionally associated with lower level cognitive functions
(Fig.~\ref{fig:networks}). Finally, we examined the most dominant components of
the brain activity patterns from each experimental condition. We used a reverse
inference approach~\citep{RubiEtal17} to identify the terms in the neuroimaging
literature most commonly associated with the corresponding maps. As summarized
in Figure~\ref{fig:neurosynth-summary}, we found that terms associated with memory and
sensory processing were associated with the strongest components in all three
story listening conditions. Terms associated with sensory integration were
associated with the strongest components in the intact and paragraph-scrambled
conditions. Terms associated with sentence comprehension, emotion, and valence
were associated with the strongest components in the intact condition. Finally,
terms associated with the default mode network were associated with the
strongest components in the word-scrambled and resting state conditions. Taken
together, our findings indicate that the informativeness and compressibility of
our brain activity patterns are task-dependent, and these properties change
systematically with factors like cognitive richness and depth of processing.

@@

We used an across-participant correlation-based classifier to decode which
stimulus timepoint matched each timepoint's neural pattern. For a given value
of $k$ (i.e., number of principal components), we first used group PCA to
project the data from each condition into a shared $k$-dimensional space. Next,
we divided the participants into two groups: a template group,
$\mathcal{G}_{\mathrm{template}}$ (i.e., training data), and a to-be-decoded
group, $\mathcal{G}_{\mathrm{decode}}$ (i.e., test data). We averaged the
projected data within each group to obtain a single $T$ by $k$ matrix for each
group. Next, we correlated the rows of the two averaged matrices to form a $T$
by $T$ decoding matrix, $\mathbf{\Lambda}$. In this way, the rows of
$\mathbf{\Lambda}$ reflected timepoints from the template group, while the
columns reflected timepoints from the to-be-decoded group. We used
$\mathbf{\Lambda}$ to assign temporal labels to each timepoint (row) from the
test group's matrix, using the row of the training group's matrix with which it
was most highly correlated. We repeated this decoding procedure, but using
$\mathcal{G}_{\mathrm{decode}}$ as the template group and
$\mathcal{G}_{\mathrm{template}}$ as the to-be-decoded group. Given the true
timepoint labels (for each group), we defined the decoding accuracy as the
average proportion of correctly decoded timepoints, across both groups (where
chance perfomance is $\frac{1}{T}$). In Figures~\ref{fig:inflection}
and~\ref{fig:inflection-quarters} we report the decoding accuracy for each
condition and value of $k$, averaged across $n = 100$ cross validation folds.

\subsection*{Reverse inference}

To help interpret the brain activity patterns we found within the contexts of
other studies, we created summary maps of each principal component, for each
experimental condition. Each principal component comprises 700 ``weights'' on
each of the HTFA-derived RBF nodes (see \textit{Hierarchical Topographic Factor
Analysis}). For each node, we evaluated its RBF at the locations of every voxel
in the standard 2~mm MNI152 template brain and multiplied the RBF by the node's
weight. The sum of these weighted RBF activation maps provides a full-brain
image, in MNI152 space, of the given principal component.

Next, we considered 80 topics estimated using Latent Dirichlet
Allocation~\citep{BleiEtal03} applied to 9,204 functional neuroimaging articles
in the Neurosynth database~\citep{RubiEtal17}. The topics, as well as
associated brain maps identified using Neurosynth, were identified and reported
in several prior studies~\citep{FoxEtal14, SulEtal17, ChenEtal20}. The topic
labels for each topic were generated automatically with the following ChatGPT
(chat.openai.com) prompt: ``Please help me come up with intuitive labels for
topic topics I found by fitting a topic model to thousands of neuroscience and
psychology articles. I'll paste in the top 10 highest-weighted words for each
topic, and I'd like you to respond with a suggested label. For each topic,
please respond with just the topic label and no other formatting or text. Here
are the next topic's top words:'' followed by a comma-separated list of the
given topic's top-weighted words reflected in the Table~\topics. For some
topics, ChatGPT responded with a longer-form response rather than a concise
topic label. In these instances, on a case-by-case basis, we used a second
follow-up prompt to achieve the given topic's label: ``could you please come up
with a more concise label for that topic?''. We then manually identified a set
of 11 cognitive labels that were intended to encapsulate a representative range
of widely studied low-level and high-level cognitive functions. In choosing the
set of cognitive labels, we jointly considered each topic's ChatGPT-derived
topic label, along with the top-weighted words for the topic. We attempted to
generate a concise set of labels that still spanned the full set of cognitive
functions reflected across the 80 topics. Topics that appeared unrelated to
specific cognitive functions (e.g., topics related to specific methods or
clinical themes) are designated with dashes in Table~\topics.

Finally, following an approach used in several prior studies~\citep{FoxEtal14,
SulEtal17, ChenEtal20} we treated the correlation between a given component's
brain map and each topic's brain map as a reflection of how much the component
was reflective of the given topic. This resulted in a set of 80 ``weights''
(correlation coefficients) for each component's brain map, with one weight per
Neurosynth-derived topic.

@@

We manually identified 11 cognitive labels
spanning the set of 80 Neurosynth-derived topics: cognitive control, language
processing, memory, emotion, social cognition, spatial cognition, attention,
reward, sensory perception, motor control, and resting state. We then used
ChatGPT to automatically ``rank'' the processes from high-level to low-level
using the following prompt: ``Please rank these cognitive processes from
highest-level to lowest-level, where higher values indicate higher-order or
higher-level processes. return the result as a csv file with a header row and
two columns: `Cognitive label' and `Rank'. Here are the processes: cognitive
control, language processing, memory, emotion, social cognition, spatial
cognition, attention, reward, sensory perception, motor control, resting
state''.  Table~\topicTags displays the output.  We used these labels in
the analysis presented in Figure~\ref{fig:neurosynth-summary}E to help summarize
difference in topic weightings across experimental conditions.

### 2023-03-16 paper/supplement.tex
checking in edits

\documentclass[english]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{apacite}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{pxfonts}
\usepackage[utf8]{inputenc}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage[left]{lineno}
\usepackage{soul}
\usepackage[table]{xcolor}
\usepackage{colortbl}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{lscape}
\usepackage{rotating}
\usepackage{xtab}

@@

\begin{landscape}
  \tablehead{\toprule
  Topic label &     Cognitive label &           Term 1 &        Term 2 &          Term 3 &         Term 4 &      Term 5 &         Term 6 &        Term 7 &         Term 8 &        Term 9 &        Term 10 \\
  \midrule
  }
  \tabletail{\bottomrule}
  \tiny

\begin{xtabular}{rlllllllllll}

@@

 \label{tab:topics}}
\end{xtabular}

\caption{\vspace{0.2cm} \normalsize \noindent Table S1: \textbf{Neurosynth-derived topics.} We report the top-weighted terms
for each of 80 topics identified using Latent Dirichlet
Allocation~\citep{BleiEtal03} applied to 9,204 functional neuroimaging articles
in the Neurosynth database~\citep{RubiEtal17}. The topics, as well as
associated brain maps identified using Neurosynth, were identified and reported
in several prior studies~\citep{FoxEtal14, SulEtal17, ChenEtal20}. The topic
labels for each topic were generated automatically with the following ChatGPT
(chat.openai.com) prompt: ``Please help me come up with intuitive labels for
topic topics I found by fitting a topic model to thousands of neuroscience and
psychology articles. I'll paste in the top 10 highest-weighted words for each
topic, and I'd like you to respond with a suggested label. For each topic,
please respond with just the topic label and no other formatting or text. Here
are the next topic's top words:'' followed by a comma-separated list of the
given topic's top-weighted words reflected in the table. For some topics,
ChatGPT responded with a longer-form response rather than a concise topic
label. In these instances, on a case-by-case basis, we used a second follow-up
prompt to achieve the given topic's label: ``could you please come up with a
more concise label for that topic?''. We then manually identified a set of 11
cognitive labels that were intended to encapsulate a representative range of
widely studied low-level and high-level cognitive functions. In choosing the
set of cognitive labels, we jointly considered each topic's ChatGPT-derived
topic label, along with the top-weighted words for the topic. We attempted to
generate a concise set of labels that still spanned the full set of cognitive
functions reflected across the 80 topics. Topics that appeared unrelated to
specific cognitive functions (e.g., topics related to specific methods or
clinical themes) are designated with dashes.

\end{landscape}

@@

\caption{\textbf{Top terms associated with the highest-weighted components by
condition, broken down by story segment.} Each group of five rows corresponds
to an experimental condition (denoted by color, as indicated in the legend in
the lower right), and the columns and shading correspond to the component
number (ranked by proportion of variance explained). The colored squares in
front of many of the topics denote manually identified cognitive labels
(Tab.~S1).}

@@

\caption{\textbf{Functions associated with top-weighted components by
condition.} \textbf{A. Top-weighted topics by condition.} Here we display
per-condition (rows, indicated by colored dots) topic correlations, averaged
across topics that pertain to each of several broad cognitive functions
(columns within each sub-panel, indicated by colored dots). Each sub-panel
reflects correlations for the components indicated in the panel titles. A
legend for the condition and cognitive function classifications is displayed in
the lower right of the figure. Table~S1 provides a list of each topic's
top-weighted terms, along with each topic's manually labeled cognitive
classification. A full list of the topics most highly associated with each
component may be found in Figure~\ref{fig:top-terms}. \textbf{B. Associations
between per-condition components and cognitive functions.} The network plots
denote positive average correlations between the component images for each
condition (gray-outlined dots on the left sides of each network; colors denote
conditions) and topic-specific brain maps associated with each indicated
cognitive function (black-outlined dots on the right sides of each network;
colors denote cognitive functions). The line thicknesses are proportional to
the correlation values (correlation coefficients are noted in the heatmaps in
Panel A). \textbf{C. Correlations between each principal component, by
condition.} The heatmaps display the correlations between the brain maps
(Fig.~\componentBrains) for each principal component (sub-panel), across each
pair of conditions (rows and columns of each sub-panel's matrix, indicated by
colored dots). \textbf{D. Associations between per-condition topic weights, by
component.} Each sub-panels network plot summarizes the pattern of correlations
between the topic correlations from each of the $n$\textsuperscript{th} top-weighted principal components
(sub-panel), for each experimental condition (gray-outlined dots). The line
thicknesses are proportional to the correlation values (correlation
coefficients are noted in the heatmaps in Panel C).}

>>>>

\documentclass[english]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{apacite}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{pxfonts}
\usepackage[utf8]{inputenc}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage[left]{lineno}
\usepackage{soul}
\usepackage{booktabs}
\usepackage{supertabular}
\usepackage{lscape}

@@

%\begin{landscape}

  \begin{landscape}
  \begin{center}
\tiny
  \tablehead{
  \toprule
  Topic label &     Cognitive label &           Term 1 &        Term 2 &          Term 3 &         Term 4 &      Term 5 &         Term 6 &        Term 7 &         Term 8 &        Term 9 &        Term 10 \\
  \midrule}
  %\midrule
  %}
  \tabletail{\bottomrule}

  \bottomcaption{\textbf{Neurosynth-derived topics.} We report the top-weighted terms for each
  of 80 topics identified using Latent Dirichlet Allocation~\citep{BleiEtal03}
  applied to 9,204 functional neuroimaging articles in the Neurosynth
  database~\citep{RubiEtal17}. See \textit{Reverse inference} for additional
  information.}
%\begin{xtabular}{rlllllllllll}

  \begin{supertabular}{rlllllllllll}

@@

 \label{tab:topics}

\end{supertabular}
\end{center}
\end{landscape}

\newpage

\begin{table}[t]
\begin{center}
\begin{tabular}{lr}
  \toprule
  Cognitive label &  Rank \\
  \midrule
  Cognitive control   &    10 \\
  Language processing &     9 \\
  Memory              &     8 \\
  Emotion             &     7 \\
  Social cognition    &     6 \\
  Spatial cognition   &     5 \\
  Attention           &     4 \\
  Reward              &     3 \\
  Sensory perception  &     2 \\
  Motor control       &     1 \\
  Resting state       &     0 \\
  \bottomrule
\end{tabular}

\caption{\textbf{Ranking cognitive processes.} The table displays the output of
a ChatGPT~\citep{ChatGPT} prompt asking for a ranking of the cognitive
processes reflected in the labels from Table~\ref{tab:topics}. See \textit{Ranking cognitive
processes} for additional detail.}
\label{tab:topic-tags}

\end{center}
\end{table}

\newpage

@@

\caption{\textbf{Top terms associated with the highest-weighted components by
condition, broken down by story segment.} Each group of five rows corresponds
to an experimental condition (denoted by color, as indicated in the legend in
the lower right), and the columns and shading correspond to the component
number (ranked by proportion of variance explained). The colored squares in
front of many of the topics denote manually identified cognitive labels
(Tab.~\ref{tab:topics}).}

@@

\caption{\textbf{Functions associated with top-weighted components by
condition.} \textbf{A. Top-weighted topics by condition.} Here we display
per-condition (rows, indicated by colored dots) topic correlations, averaged
across topics that pertain to each of several broad cognitive functions
(columns within each sub-panel, indicated by colored dots). Each sub-panel
reflects correlations for the components indicated in the panel titles. A
legend for the condition and cognitive function classifications is displayed in
the lower right of the figure. Table~\ref{tab:topics} provides a list of each topic's
top-weighted terms, along with each topic's manually labeled cognitive
classification. A full list of the topics most highly associated with each
component may be found in Figure~\ref{fig:top-terms}. \textbf{B. Associations
between per-condition components and cognitive functions.} The network plots
denote positive average correlations between the component images for each
condition (gray-outlined dots on the left sides of each network; colors denote
conditions) and topic-specific brain maps associated with each indicated
cognitive function (black-outlined dots on the right sides of each network;
colors denote cognitive functions). The line thicknesses are proportional to
the correlation values (correlation coefficients are noted in the heatmaps in
Panel A). \textbf{C. Correlations between each principal component, by
condition.} The heatmaps display the correlations between the brain maps
(Fig.~\ref{fig:components}) for each principal component (sub-panel), across each
pair of conditions (rows and columns of each sub-panel's matrix, indicated by
colored dots). \textbf{D. Associations between per-condition topic weights, by
component.} Each sub-panels network plot summarizes the pattern of correlations
between the topic correlations from each of the $n$\textsuperscript{th}
top-weighted principal components (sub-panel), for each experimental condition
(gray-outlined dots). The line thicknesses are proportional to the correlation
values (correlation coefficients are noted in the heatmaps in Panel C).}
